Geometric models reveal behavioral and neural signatures of how naturalistic experiences are transformed into episodic memories

### 2020-08-17 paper/main.tex

% JRM STOPPED HERE...

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the video (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the individual topics' overall timecourses are not as cleanly delineated as the video topics'.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall topic trajectory (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we leveraged an HMM-based optimization procedure (see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While, some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appear to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original video, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed somewhat evenly throughout the video.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the video. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the video.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the video.  All panels: error bars denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

@@

Statistical models of memory studies often treat recall success as binary (i.e., an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, the feature-rich content of a naturalistic experiences may later be described with many, highly variable levels of success.  Our framework produces a content-based model of individual stimulus and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  We define precision as the ``completeness" of recall, or how fully the presented content was recapitulated in memory.  Under our framework, we quantify this for a given recall event as the correlation between the topic proportions of the recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which we define as the ``specificity" of recall, or how unique the description of a given section of content was, compared to descriptions for other sections of content.  We quantify this for each recall event as 1 minus the average correlation between the given recall event and all other recall events not matched to the same video event.  In addition to individual events, one may also use these metrics to describe each participant's overall performance (i.e., by averaging across a participant's event-wise precision or distinctiveness scores).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, a higher precision score is correlated to both hand-annotated memory performance (Pearson's $r(15) = 0.60, p = 0.010$) and the number of video events successfully remembered, as determined by our model (Pearson's $r(15) = 0.90, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that this distinctiveness score is related to our model's estimated number of recalled events (Pearson's $r(15) = 0.55, p = 0.021$), and while we do not find distinctiveness to be related to hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$), this is not entirely surprising given how the hand-annotated memory scores were computed (see \textit{Methods} and \textit{Discussion} for details).

>>>>

Although the example participant's recall topic proportions matrix has some visual similarity to the video topic proportions matrix, the time-varying topic proportions for the example participant's recalls are not as sparse as those for the video (compare Figs.~\ref{fig:model}A and D).  Similarly, although there do appear to be periods of stability in the recall topic dynamics (i.e., most topics are active or inactive over contiguous blocks of time), the changes in topic activations that define event boundaries appear less clearly delineated in participants' recalls than in the episode's annotations.  To examine these patterns in detail, we computed the timepoint-by-timepoint correlation matrix for the example participant's recall trajectory (Fig.~\ref{fig:model}E).  As in the video correlation matrix (Fig.~\ref{fig:model}B), the example participant's recall correlation matrix has a strong block diagonal structure, indicating that their recalls are discretized into separated events.  As for the video correlation matrix, we leveraged an HMM-based optimization procedure (see \textit{Methods}) to determine how many events are reflected in the participant's recalls and where specifically the event boundaries fall (outlined in yellow).  We carried out a similar analysis on all 17 participants' recall topic proportions matrices (Fig.~\corrmats).

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  While some participants' recall topic proportions segmented into just a few events (e.g., Participants P4, P5, and P7), others' segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory, participants appear to be integrating across multiple events, blending elements of previously recalled and not-yet-recalled content into each newly recalled event~\citep[Figs.~\ref{fig:model}E, \corrmats; also see][]{MannEtal11, HowaEtal12, Mann19}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events via HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented" events from the original video, and a (potentially differently ordered) sequence of ``recalled" events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{AtkiShif68, PostPhil65, WelcBurn24}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}B). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed approximately evenly throughout the video.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=1\textwidth]{figs/list_learning}
  \caption{\small \textbf{Naturalistic extensions of classic list-learning memory analyses.} \textbf{A.} The probability of first recall as a function of the serial position of the event in the video. \textbf{B}.  The probability of recalling each event, conditioned on having most recently recalled the event \textit{lag} events away in the video.  \textbf{C.} The proportion of participants who recalled each event, as a function of the serial position of the events in the video.  All panels: error ribbons denote bootstrap-estimated standard error of the mean.}
  \label{fig:list-learning}
\end{figure}

@@

Statistical models of memory studies often treat recall success as binary (i.e., an item either was or was not recalled), or occasionally categorical \citep[e.g., to distinguish familiarity from recollection;][]{YoneEtal02}.  Such approaches are tenable in classical list-learning or recognition memory paradigms, as the presented stimuli tend to be very simple (e.g., a sequence of individual words or items).  However, memory for naturalistic experiences is much more nuanced.  For example, certain aspects of an experience might be correctly remembered at varying levels of detail, distorted, or forgotten.  Further, each remembering is itself a richly structured phenomenon.  Our framework produces a content-based model of individual video and recall events by projecting the dynamic content of the video and participants' recalls into a shared topic space.  This allows for direct, quantitative comparisons between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel, \textit{continuous} metrics for analyzing naturalistic memory:~\textit{precision} and \textit{distinctiveness}.  Precision is intended to capture the ``completeness'' of recall, or how fully the presented content was recapitulated in memory.  We define a recall event's precision as the maximum correlation between the topic proportions of that recall event and any video event (Fig.~\ref{fig:precision-distinctiveness}).  A second novel metric we introduce here is \textit{distinctiveness}, which is intended to capture the ``specificity'' of recall.  In other words, distinctiveness quantifies how much the transcript of a given recalled event references that event in particular, as opposed to other video events.  To compute a recall event's distinctiveness, we first identify the video event that it's topic vector is most strongly correlated with.  We then define distinctiveness as 1 minus the average correlation between the given recall event and all \textit{other} video events.  In addition to individual events, one may also use these metrics to describe each participant's overall performance (i.e., by averaging across a participant's event-wise precision or distinctiveness scores).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that, across participants, higher precision scores are positively correlated with both hand-annotated memory performance (Pearson's $r(15) = 0.60, p = 0.010$) and the number of video events successfully remembered, as determined by our model (Pearson's $r(15) = 0.90, p < 0.001$).  We also hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  We find that participants' distinctiveness scores were
correlated with our model's estimated number of recall events (Pearson's $r(15) = 0.55, p = 0.021$).  However, we found no evidence that distinctiveness scores were correlated with hand-annotated memory performance (Pearson's $r(15) = 0.28, p = 0.275$).  We elaborate on this potential discrepency in the \textit{Discussion} section.

%% JRM STOPPED HERE
